---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# logtree

<!-- badges: start -->
<!-- badges: end -->

logtree renders nested process execution as a live, colored tree in the
console -- tree connectors, status glyphs, and elapsed time per step --
while keeping nesting depth correct even when a step errors partway
through.

```
> Loading pipeline configuration
  |- i Reading config.yml
  |- + Validated 12 parameters              0.03s
  |- + Done                                  0.15s
> Fetching articles
  |- i Connecting to API
  |- ! Retry 1/3 due to timeout
  |- + Fetched 1,204 articles                4.2s
  |- + Done                                  4.4s
x Classification failed
  |- Error: model timeout after 30s
```

## Installation

You can install the development version of logtree from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("<github-user>/logtree")
```

## Example

```{r example}
library(logtree)

pipeline <- function() {
  log_step("Pipeline")
  load_config()
}

load_config <- function() {
  log_step("Load config")
  log_info("Reading config.yml")
  log_success("Validated 12 parameters")
}

pipeline()
```

Steps auto-close when the function that opened them returns -- normally,
early, or via an uncaught error -- so nesting depth never gets stuck out of
sync. See `vignette("logtree")` for error handling with `with_logging()`,
themes, output sinks, and verbosity control.
