---
title: "Get started"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(logtree)
```

This guide covers every feature of logtree, one section each, with a runnable
example and its output. Sections are independent -- read the first two, then
jump to whatever you need.

Each section ends with the reference pages for the functions it used and, where
one exists, a link to a complete end-to-end run in the
[Examples](https://IvanSortino.github.io/logtree/articles/examples.html)
gallery.

## The model

logtree has exactly two kinds of line.

A **step** is a node: a unit of work with a beginning, an end, an outcome, and
a duration. It prints an open line when it starts and a close line when it
finishes, and anything logged in between is nested underneath it.

A **leaf** is a message logged against whichever step is currently open. It has
a status -- debug, info, success, warning, error -- but no duration and no
children.

That is the whole vocabulary. Depth comes from R's own call stack rather than
from anything you pass: a step opened inside another step is one level deeper,
because the function that opened it was called from inside the function that
opened the outer one.

```{r}
logtree_reset()

load_config <- function() {
  log_step("Load config")
  log_info("reading config.yml")
  log_success("validated 12 parameters")
}

pipeline <- function() {
  log_step("Pipeline")
  load_config()
}

pipeline()
```

`load_config()` does not know it is being called from inside a step, and
`pipeline()` does not know that `load_config()` logs. Neither passes the other
a depth, an id, or a handle. That independence is the point: instrumented
functions compose without coordinating.

A single line is built from up to six columns, four of them optional:

<p align="center">
<img src="concept-anatomy.svg" alt="One logtree line with its timestamp, rails, connector, glyph, message and call-site columns labelled" width="700" />
</p>

The rails and connector come from depth, the glyph from status, and the three
opt-in columns -- timestamp, elapsed time, call site -- are theme slots covered
later on. Every glyph declares its own display width, which is what keeps the
message column aligned when you swap themes.

**Reference:** `log_step()`, `log_info()`, `logtree_reset()`

## Steps that close themselves

`log_step()` opens a step and registers its close on **the calling function's
frame**, not on its own. The close therefore fires when that function exits --
by returning normally, by an early `return()`, or because an error unwound
through it.

```{r}
logtree_reset()

validate <- function(rows) {
  log_step("Validate")
  if (rows == 0) {
    log_warn("nothing to validate")
    return(invisible(NULL))       # early return: the step still closes
  }
  log_success("all rows valid")
}

check <- function() {
  log_step("Check")
  validate(0)
  validate(12)
}

check()
```

Both `validate()` calls closed at the right depth, and the second one is a
sibling of the first rather than a child, even though the first left through a
`return()` in the middle of its body. There is no path out of a function that
skips the close, which is what stops indentation from drifting during a long
run.

The mechanism is invisible in the output, so it is worth seeing laid out
against a run. The bars are how long each frame lives:

<p align="center">
<img src="concept-frames.svg" alt="An execution trace beside the logtree output it produces, with bars showing each function frame's lifetime" width="700" />
</p>

Notice the two rows that print a close line: neither of them is a `log_*()`
call. They are returns. The close is written by the frame ending, which is why
it cannot be skipped and why nothing has to remember to balance it.

### At the top level

`log_step()` needs a frame to attach to. At the top level of a script or
console session there is not one -- the global environment never "returns" --
so the step would stay open forever. logtree prints a one-time nudge if you do
this, and points you at the manual pair instead:

```r
log_open("Section")   # opens
# ... work ...
log_close()           # closes
```

See [Manual step control](#manual-step-control).

**Reference:** `log_step()`, `log_open()`, `log_close()` &middot;
**Example:** [Nightly ETL](https://IvanSortino.github.io/logtree/articles/examples.html#nightly-etl)

## Leaf lines and levels

Five functions log a leaf under the current step. They differ in status, in
what they mean, and in whether they touch the step around them.

| Function | Level | Elevates its step? | For |
| --- | --- | --- | --- |
| `log_debug()` | most verbose | no | diagnostic detail, hidden by default |
| `log_info()` | ordinary | no | what is happening |
| `log_success()` | ordinary | no | a milestone reached |
| `log_warn()` | high | yes | something survivable went wrong |
| `log_error()` | highest | yes | something failed, without throwing |

```{r}
logtree_reset()

fetch <- function() {
  log_step("Fetch")
  log_debug("cache miss for key user:42")
  log_info("requesting from API")
  log_warn("rate limit at 80%")
  log_success("fetched 128 rows")
}

fetch()
```

The debug line is missing because the default threshold is `"info"`; see
[Verbosity](#verbosity).

All five take the same two extra arguments: `close = TRUE` closes the enclosing
step immediately after logging the line, and `summary = TRUE`/`FALSE` pins the
line into the run digest or keeps it out regardless of its status (see
[The run digest](#the-run-digest)).

**Reference:** `log_info()`, `log_success()`, `log_warn()`, `log_error()`,
`log_debug()`

## Status elevation

`log_warn()` and `log_error()` do something the other three do not: they raise
the status of the nearest open step, so its close line reports the outcome even
though the function that opened it returned normally.

Statuses are ordered `running < success < warning < error`, and elevation only
ever moves up that order. One warning among fifty successes leaves the step
marked as a warning; a later success cannot quietly clear it.

```{r}
logtree_reset()

parse_rows <- function() {
  log_step("Parse rows")
  log_info("1,200 rows")
  log_warn("coerced 3 rows to NA")
  log_success("parsed")            # does not undo the warning
}

parse_rows()
```

<p align="center">
<img src="concept-elevation.svg" alt="A logtree tree where a warning leaf elevates its enclosing step's close glyph" width="720" />
</p>

Nothing was thrown here. `log_error()` behaves the same way -- it records a
failure and lets the run continue, which is what you want for an error you have
already recovered from. When you know the recovery worked, close the step
explicitly to override the elevated glyph:

```{r}
logtree_reset()

connect <- function() {
  log_step("Connect")
  log_error("primary unreachable (timeout after 5s)")
  log_info("failing over to replica")
  log_success("connected to replica")
  log_close(status = "success")    # override: we recovered
}

connect()
```

**Reference:** `log_warn()`, `log_error()`, `log_close()` &middot;
**Example:** [A recovered failure](https://IvanSortino.github.io/logtree/articles/examples.html#a-recovered-failure)

## Uncaught errors

Elevation covers errors you handle. For errors you do not, wrap the run in
`with_logging()`. It installs a calling handler that, at the moment the error is
signalled and **before** the stack unwinds, marks every currently-open step as
failed and logs the condition message as a leaf at the depth it happened. Then
it prints its run summary line and rethrows.

```{r, error = TRUE}
logtree_reset()

apply_migration <- function() {
  log_step("Apply migration")
  log_info("adding column users.tier")
  stop("constraint violation on users.email")
}

release <- function() {
  log_step("Release v2.1")
  apply_migration()
}

with_logging(release())
```

`with_logging()` never swallows an error. The `Error in ...` line above is the
original condition, rethrown after logging -- so `tryCatch()` around the whole
thing still works exactly as it would have.

Without `with_logging()`, depth tracking is still correct, because the close is
tied to the frame either way. What is lost is the diagnosis: no handler saw the
condition, so the step cannot be painted red retroactively. It closes as
**interrupted** -- a dimmed glyph meaning "this never finished" -- rather than
claiming success:

```{r}
logtree_reset()

risky <- function() {
  log_step("Risky")
  stop("boom")
}

try(risky(), silent = TRUE)
```

That distinction is worth keeping in mind when reading a saved log: a dimmed
step means the run was not wrapped, not that the failure was less serious.

`with_logging(global = TRUE)` installs the same handling at the top level of a
script, where there is no expression to wrap -- see
[Recipes](https://IvanSortino.github.io/logtree/articles/recipes.html).

**Reference:** `with_logging()` &middot;
**Example:** [A migration that fails](https://IvanSortino.github.io/logtree/articles/examples.html#a-migration-that-fails)

## Routing R conditions

R code you call is going to `warning()` and `message()` at you. By default those
go to stderr, outside the tree, and end up in a different file or nowhere at
all. `with_logging(warnings = TRUE)` routes them in: a `warning()` becomes a
`log_warn()` leaf and a `message()` becomes a `log_info()` leaf, at the depth
where it happened.

```{r}
logtree_reset()

noisy <- function() {
  log_step("Load data")
  message("using cached schema")
  warning("3 rows coerced to NA")
  log_info("1,200 rows")
}

with_logging(noisy(), summary = FALSE, warnings = TRUE)
```

Once routed they are ordinary leaves: they reach every sink, they reach the
digest, and the routed warning elevates its step exactly as `log_warn()` would.
In real terminal colour that reads as:

<p align="center">
<img src="routed-conditions.svg" alt="A logtree tree where warning() and message() have become warn and info leaves" width="1000" />
</p>

### What it costs

Routing means muffling. A routed condition stops at the leaf, so it no longer
reaches `warnings()`, your own handlers, or stderr. The tree becomes the single
record of the run -- which is the point -- but it is a trade, and a routed
warning also elevates its enclosing step, so wrapping third-party code that
warns freely will turn steps yellow.

That is why it is opt-in, and why you can name one kind and not the other:

| `warnings =` | `warning()` | `message()` |
| --- | --- | --- |
| `FALSE` (default) | stderr, untouched | stderr, untouched |
| `TRUE` | `log_warn()` leaf, muffled, elevates the step | `log_info()` leaf, muffled |
| `"warning"` | `log_warn()` leaf, muffled, elevates the step | stderr, untouched |
| `"message"` | stderr, untouched | `log_info()` leaf, muffled |

**Reference:** `with_logging()`

## Manual step control

`log_open()` opens a step and returns its id; `log_close()` closes it. They are
the pair to reach for wherever there is no function frame to hang a close on --
the top level of a script, a loop body, a block of a report.

```{r}
logtree_reset()

id <- log_open("Import")
log_info("reading three files")
log_success("9,412 rows")
log_close(id)
```

`log_close()` with no arguments closes the innermost open step, which is
usually what you want; passing an id closes that specific one (and anything
still open inside it). `status =` overrides the outcome, as in
[Status elevation](#status-elevation) above.

Two shortcuts cover the common cases. Every leaf function takes `close = TRUE`,
which logs the line and closes the step in one call:

```{r}
logtree_reset()

id <- log_open("Publish")
log_success("pushed to production", close = TRUE)
```

And `log_step()` and `log_open()` take `parent =` for the rare case where a
step belongs under something other than the innermost open node.

### Re-running the same line

At the top level, a step is keyed on its own source location. Re-running the
same `log_open()` line in RStudio or Positron therefore re-anchors to the same
node rather than nesting a level deeper on every run -- so an interactive
session where you keep re-evaluating a block does not walk off the right edge
of the console.

**Reference:** `log_open()`, `log_close()`, `log_step()`

## Grouping

Sometimes a run has twenty steps that are all the same *kind* of thing, and
nesting each on its own line buries the structure. `group =` collapses adjacent
steps that share a value under one header:

```{r}
logtree_reset()

load_file <- function(dataset, file) {
  log_step(file, group = dataset)
  log_info("reading rows")
  log_success("merged")
}

import_datasets <- function() {
  log_step("Import datasets")
  load_file("sales",   "2023.csv")
  load_file("sales",   "2024.csv")
  load_file("returns", "2024.csv")
}

import_datasets()
```

Pass a bare value to use it as both the match key and the header, or
`c(name = value)` to show a fixed `name` while grouping on `value`.

The rule that catches people out is adjacency -- a value that comes back later
opens a *second* header rather than rejoining the first:

<p align="center">
<img src="concept-grouping.svg" alt="A logtree tree showing adjacent steps grouped under one header and a non-adjacent recurrence opening a second" width="720" />
</p>

Three properties are worth knowing:

- **Grouping is adjacency-based.** A group stays open across calls with a
  matching value and closes as soon as something else appears at its level --
  a step with a different value, a plain ungrouped step, or a leaf. The same
  value recurring later opens a *fresh* group rather than reopening the old
  header. This keeps the tree in chronological order; a global grouping would
  have to reorder the run to be useful.
- **A group is not tied to a frame.** It lingers after its last member closes,
  waiting to see whether the next entry joins it, and is popped when the
  enclosing step finishes.
- **A group's status is aggregated from its members.** If one file fails, the
  group's own close line carries the error glyph, so a collapsed group never
  hides a failure inside it.

**Reference:** `log_step()`, `log_open()` &middot;
**Example:** [Importing many files](https://IvanSortino.github.io/logtree/articles/examples.html#importing-many-files)

## Verbosity

`logtree_threshold()` sets the minimum leaf level to render: `"debug"`,
`"info"` (the default), `"warn"`, or `"error"`.

```{r}
logtree_reset()

fetch_verbose <- function() {
  log_step("Fetch")
  log_debug("cache miss for key user:42")
  log_info("connecting to API")
  log_success("fetched 12 records")
}

fetch_verbose()                  # default: the debug line is hidden

logtree_threshold("debug")
fetch_verbose()                  # raised: it appears

logtree_threshold("info")
```

Two rules keep this from doing damage:

- **Step open and close lines are never gated.** Hiding them would break the
  tree structure, so a threshold only ever removes leaves.
- **Verbosity is a rendering gate, not a recording one.** A `log_warn()`
  suppressed by `logtree_threshold("error")` still elevates its step's close
  glyph and still reaches the run digest. What you chose not to print is not
  the same as what did not happen.

`logtree_threshold()` is only the *default*. Each sink can pin a level of its
own, so a debug-level log file does not drag the console down with it -- see
[Output sinks](#output-sinks).

**Reference:** `logtree_threshold()`

## The run digest

A long tree scrolls. `logtree_summary()` prints a digest of everything that
went wrong since the last `logtree_reset()` -- every warning, error, and
interrupted step, plus any line pinned with `summary = TRUE` -- each with a
breadcrumb showing where in the tree it happened.

```{r}
logtree_reset()

migrate <- function() {
  log_step("Apply migration")
  log_warn("table lock held 800ms")
  log_error("constraint violation on users.email")
}

smoke_test <- function() {
  log_step("Smoke test")
  log_success("all endpoints 200")
}

release <- function() {
  log_step("Release v2.1")
  migrate()
  smoke_test()
}

with_logging(release(), summary = FALSE)
logtree_summary()
```

Two different things are called a summary here, and it is worth keeping them
apart:

- **The run summary line** -- `Run complete in 0.15s`, or `Run failed in ...`
  -- is printed by `with_logging()` itself. Its `summary =` argument controls
  that one line, and nothing else. The example above passes `summary = FALSE`
  purely to keep the output tidy.
- **The digest** is the block above, and it only appears when you call
  `logtree_summary()`. It is never printed for you, because what belongs in it
  depends on the run: a script wants it at the end, a test wants it not at all.

The digest also survives things the tree does not. A warning hidden by
`logtree_threshold("error")` still reaches it, and so does one from a muted
run.

Three arguments shape it: `filter` restricts to given statuses, `depth` trims
each breadcrumb to its N deepest nodes (useful when the tree is deep and the
full path is noise), and `trace` pins the call-site column for this one call.

```{r}
logtree_summary(filter = "error", depth = 1)
```

**Reference:** `logtree_summary()`, `logtree_reset()` &middot;
**Example:** [A migration that fails](https://IvanSortino.github.io/logtree/articles/examples.html#a-migration-that-fails)

## Call sites

The digest says *what* went wrong. The `trace` theme slot says *where*: it
annotates lines with `file.R:line fn()`, and the location is a terminal
hyperlink, so a click opens your editor at that line.

It is off in every preset, because capturing a call site costs a frame walk per
logged line. `show = "problems"` is the usual setting -- warnings, errors, and
interrupted steps, which is where you actually want a location:

```{r}
logtree_reset()
logtree_theme(list(trace = list(show = "problems", format = "{fn}()")))

flaky <- function() {
  log_step("Parse rows")
  log_info("1,200 rows")
  log_warn("coerced 3 rows")
}

flaky()
logtree_theme("unicode")
```

`show` also takes the statuses themselves -- `"running"` for open lines, plus
`"info"`, `"debug"`, `"success"`, `"warning"`, `"error"`, `"interrupted"` -- so
`show = "error"` annotates errors and leaves tolerated warnings bare. `TRUE`
marks every line that can carry one. An ordinary close line never carries a
call site whatever you name, since its site is its own open line's, two rows up.

`format` is a template over `{fn}`, `{file}` and `{line}`, defaulting to
`"{file}:{line} {fn}()"`. The example pins `"{fn}()"` because `{file}` and
`{line}` come from R's source references, which exist only when code was parsed
with `keep.source = TRUE` -- true interactively and under
`devtools::load_all()`, but not under plain `Rscript`, and not in this
vignette's knitted chunks. Rather than print `NA`, the template drops any
whitespace-separated run whose placeholders are all unavailable, so the default
degrades to `flaky()` on its own. In a real script it reads:

```
> Load data  R/pipeline.R:12 load_data()
+- > Parse rows  R/pipeline.R:18 parse_rows()
|  +- ! coerced 3 rows  R/pipeline.R:20 parse_rows()
```

**Capturing and printing are separate stages, and only printing can be decided
afterwards** -- once a run is over its frame stack is gone. That is what
`capture` is for: `capture = TRUE` records a call site on every line whatever
`show` prints, so a tree that stays exactly as quiet as it was can still hand
locations to the digest or a JSON sink.

```r
logtree_theme(list(trace = list(show = FALSE, capture = TRUE)))
job()                            # tree unchanged, not one call site printed
logtree_summary(trace = TRUE)    # ... but the digest has them
```

Styling the column, and pinning it per sink, are covered in the
[Themes cookbook](https://IvanSortino.github.io/logtree/articles/themes.html).

**Reference:** `logtree_theme()`, `logtree_summary()`

## Timestamps

A tree says how *long* each step took, but not *when* any of it happened --
which starts to matter once a log is read after the fact rather than watched as
it runs. The `timestamp` slot puts a wall-clock column in front of every line.
It is off in every preset (`format = NULL`), so it costs nothing until asked
for.

```{r}
logtree_reset()
logtree_theme(list(timestamp = list(format = "%H:%M:%S")))
pipeline()
logtree_theme(list(timestamp = list(format = NULL)))
```

`"%H:%M:%S"` is the interactive choice; `"%Y-%m-%d %H:%M:%S"` is what a saved
log wants, since a file outlives the day it was written.

In the coloured presets the column ships `"silver"`: it is supporting detail,
so it stays faint enough that the status glyphs remain what your eye lands on.

<p align="center">
<img src="timestamp-silver.svg" alt="A logtree tree with a silver wall-clock timestamp column in front of every line" width="740" />
</p>

Every line kind takes the column at one fixed left edge, and the width is
measured from a rendered sample rather than from the format string, so a format
whose width varies with the value cannot shear the tree. The digest is never
stamped -- it replays events that already happened, so the time it was printed
would be the wrong answer.

**Reference:** `logtree_theme()`, `logtree_sink_file()`

## Themes and presets

`logtree_theme()` swaps the whole glyph and colour preset. There are five:

| Preset | What it is for |
| --- | --- |
| `"unicode"` | The default. Box-drawing connectors, coloured symbol glyphs, for an interactive terminal. |
| `"ascii"` | Plain ASCII, no colour. Safe for log files and non-UTF-8 terminals; also what every text file sink renders through. |
| `"emoji"` | Emoji status glyphs (width-2 cells) over box-drawing connectors. |
| `"minimal"` | No connectors at all -- depth is carried by indentation alone. |
| `"ci"` | Bracketed word glyphs over pure-ASCII connectors, no colour, so a failure greps as `[fail]`. |

```{r}
logtree_theme("ascii")
pipeline()

logtree_theme("ci")
pipeline()

logtree_theme("unicode")
```

Individual slots are overridden with `overrides`, a named list keyed by slot,
each element holding only the fields to change:

```{r}
logtree_theme("unicode", overrides = list(
  success = list(glyph = "*", color = c("green", "bold")),
  done    = list(text = "{label} ok")
))
pipeline()
logtree_theme("unicode")
```

Switching themes never breaks column alignment, because each glyph declares its
own rendered width rather than having it measured -- `nchar()` cannot reliably
size an emoji cell.

The full list of slots and fields, the elapsed-time controls, close-line text
templating, and recipes for building your own preset are in the
[Themes cookbook](https://IvanSortino.github.io/logtree/articles/themes.html).

**Reference:** `logtree_theme()`

## Layout and density

Four arguments control horizontal space, independently of any glyph. They are
scalars carried on the theme, set through `logtree_theme()`'s own arguments
rather than through `overrides`, and cleared by the next preset swap.

| Argument | The gap it sets |
| --- | --- |
| `compact` | the per-level tree column: `"medium"` drops the trailing gap after each connector, `"tight"` also slims the connectors to one character |
| `connector_gap` | a leaf or close line's own connector, to its status glyph |
| `glyph_gap` | the status glyph, to the message text |
| `wrap` | the column budget a rendered line is capped at |

```{r}
logtree_theme("unicode", compact = "tight", glyph_gap = 0)
pipeline()
logtree_theme("unicode")
```

`wrap = TRUE` follows `cli::console_width()`, measured at render time so a
terminal resized mid-run is picked up on its own; a number pins a fixed width.
Continuation lines indent to the message column and carry the rails down, so a
wrapped message still reads as one node:

```{r}
logtree_theme("unicode", wrap = 56)

long <- function() {
  log_step("Deploy")
  log_info("uploading layers to registry.example.internal, 412 MB across 14 layers")
}
long()

logtree_theme("unicode")
```

None of the four reaches file sinks, which render through the ascii preset with
its built-in spacing.

**Reference:** `logtree_theme()`

## Output sinks

Every logged event fans out to every registered sink. The console sink is
registered by default under the reserved id `"console"`; `logtree_sink_file()`
adds a file.

```{r}
log_path <- tempfile(fileext = ".log")
handle <- logtree_sink_file(log_path, format = "text")

logtree_reset()
pipeline()

writeLines(readLines(log_path))
```

`format = "json"` writes NDJSON instead -- one record per line, with the event
kind, level, depth, label, status, elapsed time, an ISO-8601 timestamp, and a
`run_id` so one run's lines can be picked out of a file many runs appended to.

Every registration returns a handle. Sinks deliberately survive
`logtree_reset()` -- they are configuration, not run state -- so
`logtree_sink_remove()` is the only way to stop one:

```{r}
logtree_sinks()
logtree_sink_remove(handle)
logtree_sinks()
```

`logtree_sink()` registers a sink of your own: any function of one argument,
called with each event. A sink that throws is skipped rather than allowed to
break the fanout -- the remaining sinks still run, and a warning naming the
offender is raised once.

```{r}
kinds <- character(0)
h <- logtree_sink(function(event) kinds <<- c(kinds, event$kind))

logtree_reset()
pipeline()
table(kinds)

logtree_sink_remove(h)
```

Each sink takes its own `threshold =`, defaulting to the global
`logtree_threshold()` read afresh per event. This is how a log file captures
debug detail while the console stays at `"info"`:

```{r}
verbose_path <- tempfile(fileext = ".log")
h <- logtree_sink_file(verbose_path, format = "text", threshold = "debug")

logtree_reset()
fetch_verbose()          # console: no debug line

writeLines(readLines(verbose_path))   # file: it is there
logtree_sink_remove(h)
```

File sinks also take `trace =` and `timestamp =`, pinning those columns
independently of the console's.

**Reference:** `logtree_sink_file()`, `logtree_sink()`, `logtree_sinks()`,
`logtree_sink_remove()` &middot;
**Examples:** [A CI build log](https://IvanSortino.github.io/logtree/articles/examples.html#a-ci-build-log),
[Structured NDJSON](https://IvanSortino.github.io/logtree/articles/examples.html#structured-ndjson)

## Testing your logging

If your package logs with logtree, you will eventually want to assert that a
pipeline logged what it should. Capturing console output and pattern-matching
glyphs and connectors is the wrong tool -- it breaks when the theme changes and
it tests the renderer rather than your code.

`logtree_sink_memory()` collects events in a capped buffer, and
`logtree_sink_memory_events()` reads them back as a data frame, one row per
event, with the same columns a JSON sink writes:

```{r}
h <- logtree_sink_memory()

logtree_reset()
pipeline()

events <- logtree_sink_memory_events(h)
events[, c("level", "depth", "label", "status")]

logtree_sink_remove(h)
```

Note the column names, since they are easy to guess wrong: `level` is the kind
of event (`"open"`, `"leaf"`, `"close"`, `"group"`, `"group_close"`) and
`status` is its outcome (`"step"`, `"info"`, `"success"`, `"warning"`, ...).
The full set is `ts`, `run_id`, `level`, `id`, `parent_id`, `depth`, `label`,
`elapsed`, `status`, `fn`, `file`, `line`. A sink function registered with
`logtree_sink()` receives a different shape -- an event *list*, whose kind is
`event$kind`.

Both views are built from one shared record, so a run replayed from a log file
and the same run read from memory cannot disagree.

**Reference:** `logtree_sink_memory()`, `logtree_sink_memory_events()` &middot;
**Example:** [Asserting in tests](https://IvanSortino.github.io/logtree/articles/examples.html#asserting-in-tests)

## Silence

`logtree_mute()` stops every sink receiving events without unregistering any of
them -- what a library that logs with logtree reaches for to keep its own test
suite quiet. `logtree_unmute()` turns it back on, and both return the state they
replaced, so a caller can restore what it found.

```{r}
was <- logtree_mute()

logtree_reset()
pipeline()          # prints nothing

logtree_unmute()
length(logtree_summary())
```

A muted run is still *recorded*: the digest can still report what went wrong,
and step bookkeeping is untouched, so depth is right the moment output comes
back.

**Reference:** `logtree_mute()`

## logger integration

If your codebase already uses the CRAN [logger](https://daroczig.github.io/logger/)
package, `logtree_logger()` routes those calls through logtree without
rewriting any of them. Call it once near the top of your script: it registers
logtree's layout, pairs it with `logger::appender_void` so logtree does the
rendering, and opens logger's own threshold so `logtree_threshold()` becomes the
single gate.

```{r, eval = rlang::is_installed("logger", version = "0.3.0")}
logtree_reset()
logtree_threshold("debug")

ns <- "my_app"
logtree_logger(namespace = ns)

process_data <- function() {
  log_step("Processing data")
  logger::log_info("reading input file", namespace = ns)
  logger::log_debug("parsed 5,000 rows", namespace = ns)
  logger::log_success("transformation complete", namespace = ns)
}

process_data()
logtree_threshold("info")
```

Severities map onto leaf levels: `FATAL`/`ERROR` become `log_error()`, `WARN`
becomes `log_warn()`, `SUCCESS` becomes `log_success()`, `INFO` becomes
`log_info()`, and `DEBUG`/`TRACE` both become `log_debug()` -- logger has two
debug-ish tiers, logtree has one.

**Reference:** `logtree_logger()`, `layout_logtree()` &middot;
**Example:** [Bridging logger](https://IvanSortino.github.io/logtree/articles/examples.html#bridging-logger)

## Where to go next

- [Examples](https://IvanSortino.github.io/logtree/articles/examples.html) --
  complete end-to-end runs.
- [Themes cookbook](https://IvanSortino.github.io/logtree/articles/themes.html)
  -- every slot and field, and recipes for your own preset.
- [Recipes](https://IvanSortino.github.io/logtree/articles/recipes.html) --
  top-level scripts, library authors, scheduled jobs.
- [Design philosophy](https://IvanSortino.github.io/logtree/articles/design.html)
  -- why depth is tied to frames, and why the corner connector only ever
  appears on a close line.
