---
title: "logtree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{logtree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(logtree)
```

logtree renders nested process execution as a live, colored tree in the
console: tree connectors, status glyphs, and elapsed time per step. Nesting
depth is tracked via frame-exit handlers, so it never desynchronizes, even
when a step errors partway through.

## Basic nesting

`log_step()` opens a step and automatically closes it when the calling
function returns -- normally, via an early `return()`, or because an error
propagated through it. Nesting composes for free: a step opened inside
another step is automatically indented one level deeper, with no
coordination needed between the two functions.

```{r}
logtree_reset()

load_config <- function() {
  log_step("Load config")
  log_info("Reading config.yml")
  log_success("Validated 12 parameters")
}

pipeline <- function() {
  log_step("Pipeline")
  load_config()
}

pipeline()
```

## Leaf lines and status elevation

`log_info()` and `log_success()` are plain leaf lines. `log_warn()` and
`log_error()` additionally elevate the *enclosing* step's status, so its
close line renders the elevated glyph even though the function that opened
it returns normally.

```{r}
logtree_reset()

fetch_articles <- function() {
  log_step("Fetching articles")
  log_info("Connecting to API")
  log_warn("Retry 1/3 due to timeout")
  log_success("Fetched 1,204 articles")
}

fetch_articles()
```

## Handling uncaught errors with `with_logging()`

If a step's code actually throws, wrap the run in `with_logging()`. Every
step still open at the moment of the error is flagged as failed before the
stack unwinds, the error is logged as a leaf line, a run summary prints,
and the error is rethrown -- `with_logging()` never silently swallows it.

```{r}
logtree_reset()

run_classifier <- function() {
  with_logging({
    log_step("Classifying")
    stop("model timeout after 30s")
  })
}

tryCatch(run_classifier(), error = function(e) invisible(NULL))
```

Even *without* `with_logging()`, depth tracking still unwinds correctly on
an uncaught error -- the step just renders as dimmed/interrupted rather
than retroactively painted red, since there was no handler installed
to catch the condition and flag it before the stack unwound.

```{r}
logtree_reset()

risky <- function() {
  log_step("risky")
  stop("boom")
}

tryCatch(risky(), error = function(e) invisible(NULL))
```

## Grouping

`log_step(label, group = value)` collapses adjacent steps that share the
same `value` under one synthetic header, instead of nesting each step on its
own. Here several files belong to the same dataset, so each dataset becomes a
header and its files sit underneath. The group stays open across calls with a
matching value and only closes -- as a sibling, with its own status
aggregated from its members -- once a step with a different value (or a plain
ungrouped step) appears at the same level, or the enclosing step finishes.
Grouping is strictly adjacency-based: the same value recurring later, after
something else has closed the group, opens a fresh one rather than reusing the
old header.

Pass a bare `value` to use it as both the match key and the header, or
`c(name = value)` to show a fixed `name` while still grouping on `value`.

```{r}
logtree_reset()

load_file <- function(dataset, file) {
  log_step(file, group = dataset)
  log_info("Reading rows")
  log_success("Merged into dataset")
}

import_datasets <- function() {
  log_step("Import datasets")
  load_file("sales",   "2023.csv")
  load_file("sales",   "2024.csv")
  load_file("returns", "2024.csv")
}

import_datasets()
```

## Themes

Five built-in presets are available. Every glyph is overridable, and
switching themes never breaks column alignment, since each glyph declares its
own rendered width rather than having that width measured from the string.

| Preset | What it is for |
| ------ | -------------- |
| `"unicode"` | The default. Box-drawing connectors and coloured symbol glyphs, for an interactive terminal. |
| `"ascii"` | Plain ASCII, no colour. Safe for log files, CI, and non-UTF-8 terminals; also what every file sink renders through. |
| `"emoji"` | Emoji status glyphs (width-2 cells) over box-drawing connectors. |
| `"minimal"` | No connectors at all -- depth is carried by indentation alone. |
| `"ci"` | Bracketed word glyphs over pure-ASCII connectors, no colour. |

```{r}
logtree_theme("ascii")
pipeline()

logtree_theme("unicode")
```

`"minimal"` drops the connectors entirely: `branch`, `corner` and `pipe` are
empty strings, so a level costs nothing but its indentation. It also ships a
lighter glyph vocabulary, dimmed times, and wordless close lines. The trade is
worth knowing -- `info`, `debug` and `interrupted` all render the middle dot
and are told apart by colour alone.

```{r}
logtree_theme("minimal")
pipeline()

logtree_theme("unicode")
```

`"ci"` is built for log capture rather than a terminal: bracketed word glyphs,
pure-ASCII connectors, and no colour in any slot, so a captured build log
survives a runner that strips ANSI and a failure greps as `[fail]`. The words
are different lengths on purpose; each declares its true width, so the message
column still lines up.

```{r}
logtree_theme("ci")
pipeline()

logtree_theme("unicode")
```

Individual glyph slots can be overridden with `overrides` -- a named list
keyed by slot, each element holding only the fields to change (everything else
is kept from the active theme):

```{r}
logtree_theme("unicode", overrides = list(
  success = list(glyph = "*", color = c("green", "bold")),
  group   = list(bracket = TRUE)
))
pipeline()
logtree_theme("unicode")
```

The tick on a `log_success()` line and the tick on a step's own `Done` line are
two different slots -- `success` and `done` -- that simply ship the same glyph
in every preset. Restyle one and the other stays as it was:

```{r}
logtree_theme("unicode", overrides = list(
  done = list(glyph = "=", color = "silver")
))
pipeline()
logtree_theme("unicode")
```

**Accepted slots** (valid names in an `overrides` list):

| Slot | Applies to | Fields it accepts |
| ---- | ---------- | ----------------- |
| `step` | open / running step glyph | `glyph`, `width`, `color` |
| `info` | `log_info()` leaf | `glyph`, `width`, `color` |
| `debug` | `log_debug()` leaf | `glyph`, `width`, `color` |
| `success` | `log_success()` leaf | `glyph`, `width`, `color` |
| `done` | a step's own close line on a clean close | `glyph`, `width`, `color`, `text` |
| `warning` | `log_warn()` / elevated step glyph | `glyph`, `width`, `color`, `text` |
| `error` | `log_error()` / elevated step glyph | `glyph`, `width`, `color`, `text` |
| `interrupted` | abnormal-exit (dimmed) glyph | `glyph`, `width`, `color`, `text` |
| `group` | group header marker | `glyph`, `color`, `bracket` |
| `elapsed` | the elapsed-time column printed on every close line | `show`, `min`, `color`, `slow`, `slow_color` |
| `trace` | the optional call-site column: where in your code a line came from | `show`, `format`, `color` |
| `branch` | child connector (`├─`) | `glyph`, `color` |
| `corner` | close-line connector (`└─`) | `glyph`, `color` |
| `pipe` | vertical rail (`│`) | `glyph`, `color` |
| `crumb` | `logtree_summary()` breadcrumb separator | `glyph`, `color`, `path_color` |
| `summary` | `logtree_summary()` divider above the digest | `gap`, `rule`, `line` |

**Accepted fields** (valid names inside a slot):

| Field | Type | Accepted values |
| ----- | ---- | --------------- |
| `glyph` | `character(1)` | Any string, including `""`. |
| `width` | `integer(1)` | Rendered display width of the glyph (`1` normal, `2` emoji/wide). Sets column alignment; status slots only. |
| `color` | `character` / `NULL` | One or more cli styles, or `NULL`. Named (`"red"`, `"cyan"`, ...), bright (`"br_red"`), backgrounds (`"bg_blue"`), styles (`"bold"`, `"dim"`, `"italic"`), or hex (`"#ff8800"`). A vector combines them, e.g. `c("red", "bold")`. |
| `text` | `character(1)` | Close-line status slots only (`done`, `warning`, `error`, `interrupted`). The word a close line prints before its time; `""` drops it. Expands `{label}` and `{elapsed}`. |
| `show` | `logical(1)` / `character(1)` | On `elapsed`, `FALSE` drops the time column entirely. On `trace`, `FALSE` (default) is off, `"problems"` marks only warnings/errors/interrupted steps, `TRUE` adds every open line and leaf. |
| `format` | `character(1)` | `trace` slot only. Template over `{fn}`, `{file}`, `{line}`. Default `"{fn}() {file}:{line}"`. |
| `min` | `numeric(1)` | `elapsed` slot only. Hide times below this many seconds. |
| `slow` | `numeric(1)` / `NULL` | `elapsed` slot only. Times at or over this count as slow. |
| `slow_color` | `character` / `NULL` | `elapsed` slot only. Styles a slow time in place of `color`. |
| `bracket` | `logical(1)` | `group` slot only. `TRUE` wraps the header name in `< >`. |
| `path_color` | `character` / `NULL` | `crumb` slot only. Styles the breadcrumb's path nodes. |
| `gap` | `integer(1)` | `summary` slot only. Blank lines above the digest. |
| `rule` | `logical(1)` / `character(1)` | `summary` slot only. The divider drawn above the digest. |
| `line` | `integer(1)` / `character(1)` | `summary` slot only. The rule's line type. |

### Close lines and elapsed time

The word a step prints when it closes is the `text` field of a theme slot. It
is read from the closing status's own slot, falling back to `done`'s and then
to the built-in `"Done"` -- so one call renames every close line, and a
narrower one renames only the outcomes that went wrong. Two placeholders are
expanded: `{label}` (the step's own label, or a group's name) and `{elapsed}`.

```{r}
logtree_theme("unicode", overrides = list(
  done  = list(text = "{label} ok"),
  error = list(text = "{label} FAILED")
))
pipeline()
logtree_theme("unicode")
```

`text` governs close lines only -- a `log_error()` leaf keeps its own message
-- and `text = ""` drops the word entirely, leaving the glyph and the time.

The time itself is the `elapsed` slot. `min` hides anything faster than a
threshold, which is how you silence the `0.00s` noise on trivial steps, while
`slow` and `slow_color` flag the ones worth noticing:

```{r}
logtree_theme("unicode", overrides = list(
  elapsed = list(min = 0.001, color = "silver", slow = 1, slow_color = "red")
))
pipeline()
logtree_theme("unicode")
```

`show = FALSE` drops the column outright. The `with_logging()` run-summary line
takes the slot's colouring but never its hiding rules, since
`Run complete in <time>` would read as an unfinished sentence without its time.

### Call sites

The `trace` slot annotates a line with where in your code it came from. It is
off in every preset -- capturing a call site costs a frame walk per logged line,
so you opt in and the default path pays nothing. `show = "problems"` marks only
warning and error leaves plus interrupted steps; `show = TRUE` adds every open
line and leaf.

```{r}
logtree_theme("unicode", overrides = list(
  trace = list(show = "problems", format = "{fn}()")
))

flaky <- function() {
  log_step("Parse rows")
  log_info("1200 rows")
  log_warn("coerced 3 rows")
}
flaky()
logtree_theme("unicode")
```

`format` is a template over `{fn}`, `{file}` and `{line}`; the default is
`"{fn}() {file}:{line}"`. The example pins `"{fn}()"` because `{file}` and
`{line}` come from R's source references, which exist only when code was parsed
with `keep.source = TRUE` -- the default interactively and under
`devtools::load_all()`, but not under plain `Rscript`, and not in this
vignette's knitted chunks. `{fn}` always works, and rather than print `NA` the
template drops any whitespace-separated run whose placeholders are all
unavailable, so the default degrades to `flaky()` on its own.

The column is appended to the message rather than given a column of its own, so
it wraps with the message instead of shearing the tree. It also reaches
`logtree_summary()`'s digest lines and the `fn` / `file` / `line` fields of a
`"json"` sink; `logtree_sink_file(trace = )` pins a file sink's column
independently of the console's, so a log file can record call sites while the
terminal stays quiet.

### Wrapping long messages

By default a long message runs off the right edge. `wrap` caps a rendered line
at a column budget instead: `TRUE` follows `cli::console_width()`, measured at
render time so a terminal resized mid-run is picked up on its own, and a number
pins a fixed width.

```{r}
logtree_theme("unicode", wrap = 56)

long <- function() {
  log_step("Deploy")
  log_info("uploading layers to the internal registry at registry.example.internal, 412 MB across 14 layers")
}
long()

logtree_theme("unicode")
```

Continuation lines indent to the message column and carry the rails down, so a
wrapped message still reads as one node of the tree: after a branch connector
the vertical rail continues, after a corner it does not. A token with no break
opportunity -- a long path, a URL -- is split by display width rather than left
to overflow, and a budget narrower than the tree is deep degrades to no
wrapping rather than to an unusable one-column line. Wrapping covers every line
logtree renders, including the `logtree_summary()` digest and the
`with_logging()` run-summary line.

### Density and spacing

Three arguments control horizontal space, independently of any glyph, each
governing one column of a rendered line:

| Argument | The gap it sets |
| -------- | --------------- |
| `compact` | the per-level tree column: `"medium"` drops the trailing gap after each connector, `"tight"` also slims the connectors to one character |
| `connector_gap` | a *leaf or close* line's own connector, to its status glyph |
| `glyph_gap` | the status glyph, to the message text -- `1` by default, `0` for the tightest look, `2` or more for a roomier message column |

```{r}
logtree_theme("unicode", compact = "tight", glyph_gap = 0)
pipeline()
logtree_theme("unicode")
```

`connector_gap` is unset by default and tracks `col_gap`, so every preset and
every `compact` density renders as it always has. Setting it lets the two
diverge -- which is what it is for: `compact = "tight"` keeps every rail column
flush while leaf and close glyphs still get air. It never touches a step's
*open* line or a group header, since those are rail columns rather than the
glyph's own approach, and it does not compound with depth.

```{r}
logtree_theme("unicode", compact = "tight", connector_gap = 1)
pipeline()
logtree_theme("unicode")
```

These three and `wrap` are *scalars* carried on the theme, not override slots:
they are set through `logtree_theme()`'s own arguments rather than through
`overrides`. All four ride the active theme and are cleared by the next preset
swap (`logtree_theme("unicode")`). None of them reaches file sinks, which render
through the ascii preset with its built-in spacing.

## Output sinks

The console sink is always on. `logtree_sink_file()` adds a plain-text or
NDJSON file sink; every logged event fans out to all active sinks
simultaneously.

```{r}
log_path <- tempfile(fileext = ".log")
logtree_sink_file(log_path, format = "text")

logtree_reset()
pipeline()

writeLines(readLines(log_path))
```

## Verbosity

`logtree_threshold()` sets the minimum leaf-line level to render
(`"debug"`, `"info"`, `"warn"`, `"error"`). Step open/close lines always
render regardless of verbosity, since hiding them would break the tree
structure; a suppressed `log_warn()`/`log_error()` still elevates the
enclosing step's close glyph.

```{r}
logtree_threshold("warn")
fetch_articles()
logtree_threshold("info")
```


`log_debug()` is the most verbose level, for fine-grained diagnostic detail.
It's hidden by default (verbosity is `"info"`) and only shown when
`logtree_threshold("debug")` is called. Like `log_info()` and
`log_success()`, it does not elevate the enclosing step's status.

```{r}
logtree_reset()

fetch_verbose <- function() {
  log_step("Fetching")
  log_debug("cache miss for key user:42")
  log_info("connecting to API")
  log_debug("request took 84ms")
  log_success("fetched 12 records")
}

# At default verbosity, debug lines are hidden
fetch_verbose()

# Raise verbosity to show debug lines
logtree_threshold("debug")
logtree_reset()
fetch_verbose()

logtree_threshold("info")
```


## Integrating with the `logger` package

If your codebase already uses the CRAN `logger` package
(<https://daroczig.github.io/logger/>) for logging, you can route those calls
through `logtree` without rewriting them. Call `logtree_logger()` once near the
top of your script: it registers `logtree`'s custom layout, pairs it with
`logger::appender_void` (a no-op appender) so `logtree` does the rendering, and
opens `logger`'s own threshold so that `logtree_threshold()` becomes the single
gate on what is shown.

```{r, eval = requireNamespace("logger", quietly = TRUE)}
logtree_reset()
logtree_threshold("debug")  # the only gate, now that logtree_logger() opened logger's

ns <- "my_app"
logtree_logger(namespace = ns)

process_data <- function() {
  log_step("Processing data")
  logger::log_info("reading input file", namespace = ns)
  logger::log_debug("parsed 5,000 rows", namespace = ns)
  logger::log_success("transformation complete", namespace = ns)
}

process_data()
```

`logger` severities map onto logtree leaf levels: `FATAL`/`ERROR` become
[log_error()], `WARN` becomes [log_warn()], `SUCCESS` becomes
[log_success()], `INFO` becomes [log_info()], and `DEBUG`/`TRACE` both become
[log_debug()] (logger has two debug-ish tiers, logtree has one). Note that
`logger`'s own `log_threshold()` and `logtree_threshold()` are two
independent gates applied on top of each other -- both apply simultaneously,
which is intentional.
