---
title: "Examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(logtree)
```

Complete runs rather than isolated features. Each example is self-contained --
copy it, run it, change it. The **Features shown** line under each heading links
back to the [Get started](https://IvanSortino.github.io/logtree/articles/logtree.html)
section that explains the piece in question.

## Nightly ETL

The shape most pipelines have: a top-level job, a few phases, leaves reporting
what each phase did. Nothing here is configured -- the nesting comes from the
call structure alone.

*Features shown:*
[steps](https://IvanSortino.github.io/logtree/articles/logtree.html#steps-that-close-themselves),
[leaf lines](https://IvanSortino.github.io/logtree/articles/logtree.html#leaf-lines-and-levels),
[status elevation](https://IvanSortino.github.io/logtree/articles/logtree.html#status-elevation),
[the run digest](https://IvanSortino.github.io/logtree/articles/logtree.html#the-run-digest)

```{r}
logtree_reset()

extract <- function() {
  log_step("Extract")
  log_info("connecting to warehouse")
  log_info("24,318 rows pulled")
  log_success("staged")
}

transform <- function() {
  log_step("Transform")
  log_info("normalising currencies")
  log_warn("112 rows missing fx rate, carried forward")
  log_success("24,318 rows transformed")
}

load_rows <- function() {
  log_step("Load")
  log_info("upserting into analytics.sales")
  log_success("24,318 rows written")
}

nightly_etl <- function() {
  log_step("Nightly ETL")
  extract()
  transform()
  load_rows()
}

with_logging(nightly_etl())
logtree_summary()
```

The `Transform` step closes with a warning glyph even though it returned
normally and reported success afterwards -- one `log_warn()` was enough to
elevate it, and elevation never moves back down. The digest is the part you
read when the run is forty steps long instead of three.

## A migration that fails

An error that nobody catches. `with_logging()` sees the condition before the
stack unwinds, so every step still open is marked failed and the message is
logged at the depth it happened -- then the error is rethrown unchanged.

*Features shown:*
[uncaught errors](https://IvanSortino.github.io/logtree/articles/logtree.html#uncaught-errors),
[the run digest](https://IvanSortino.github.io/logtree/articles/logtree.html#the-run-digest)

```{r, error = TRUE}
logtree_reset()

backup <- function() {
  log_step("Backup")
  log_success("snapshot analytics-2026-02-11 written")
}

apply_migration <- function() {
  log_step("Apply migration 0042")
  log_info("adding column users.tier")
  log_warn("table lock held 800ms")
  stop("constraint violation on users.email")
}

release <- function() {
  log_step("Release v2.1")
  backup()
  apply_migration()
  log_success("released")        # never reached
}

with_logging(release())
```

```{r}
logtree_summary()
```

`Backup` keeps its tick: it had already closed successfully when the migration
blew up, and the error handler only marks steps that are still open. That is
the difference between "the run failed" and "this step failed", and it is what
makes the tree readable after the fact.

Wrapping the call in `tryCatch()` works exactly as it would without logtree --
`with_logging()` rethrows rather than swallowing:

```{r}
logtree_reset()
result <- tryCatch(with_logging(release(), summary = FALSE),
                   error = function(e) conditionMessage(e))
result
```

## Importing many files

Twenty steps that are all the same kind of thing. Grouping collapses the
adjacent ones under a header instead of stacking twenty siblings.

*Features shown:*
[grouping](https://IvanSortino.github.io/logtree/articles/logtree.html#grouping)

```{r}
logtree_reset()

load_file <- function(dataset, file, rows, ok = TRUE) {
  log_step(file, group = dataset)
  log_info(sprintf("%s rows read", format(rows, big.mark = ",")))
  if (ok) log_success("merged") else log_error("schema mismatch, skipped")
}

import_all <- function() {
  log_step("Import datasets")
  load_file("sales",   "2023.csv", 18422)
  load_file("sales",   "2024.csv", 24318)
  load_file("returns", "2024.csv",  1204, ok = FALSE)
  load_file("returns", "2025.csv",   318)
  load_file("sales",   "2025.csv",  9871)   # not adjacent: a fresh group
}

with_logging(import_all())
logtree_summary()
```

Two things to notice. The `returns` group closes with an error glyph, because a
group's status is aggregated from its members -- collapsing a group never hides
a failure inside it. And the last `sales` file opens a *second* `sales` header
rather than rejoining the first: grouping is adjacency-based, so the tree stays
in the order things actually happened.

## A recovered failure

`log_error()` records a failure without throwing. When the recovery works, close
the step explicitly to say so -- the error stays on the record, but the step
reports the outcome it actually reached.

*Features shown:*
[status elevation](https://IvanSortino.github.io/logtree/articles/logtree.html#status-elevation),
[manual step control](https://IvanSortino.github.io/logtree/articles/logtree.html#manual-step-control)

```{r}
logtree_reset()

connect <- function() {
  log_step("Connect to database")
  log_error("primary unreachable (timeout after 5s)")
  log_info("failing over to replica db-2")
  log_success("connected, 12ms latency")
  log_close(status = "success")
}

query <- function() {
  log_step("Run report query")
  log_success("4,120 rows")
}

report <- function() {
  log_step("Daily report")
  connect()
  query()
}

with_logging(report())
logtree_summary()
```

The step closes green, but the digest still lists the error -- which is the
right split. The run succeeded; something went wrong on the way, and whoever
reads the log afterwards should know about the failover.

## A CI build log

A build runner strips ANSI, wraps lines, and greps the output. The `"ci"` preset
is built for exactly that: bracketed word glyphs, pure-ASCII connectors, no
colour anywhere.

*Features shown:*
[themes](https://IvanSortino.github.io/logtree/articles/logtree.html#themes-and-presets),
[output sinks](https://IvanSortino.github.io/logtree/articles/logtree.html#output-sinks)

```{r}
logtree_reset()
logtree_theme("ci")

build <- function() {
  log_step("Build")
  log_info("compiling 84 source files")
  log_warn("3 deprecation warnings")
  log_success("artifact 12.4 MB")
}

test <- function() {
  log_step("Test")
  log_info("running 312 tests")
  log_error("2 failures in test-parser.R")
}

ci <- function() {
  log_step("Pipeline")
  build()
  test()
}

with_logging(ci(), summary = FALSE)
logtree_summary()

logtree_theme("unicode")
```

A failure greps as `[fail]`, and the words are different lengths on purpose --
each declares its true width, so the message column still lines up.

To keep a copy on disk as well as on stdout, register a text file sink. Text
sinks always render through the ascii preset, so the file is stable whatever
the console is themed as:

```{r}
log_path <- tempfile(fileext = ".log")
h <- logtree_sink_file(log_path, format = "text")

logtree_reset()
with_logging(ci(), summary = FALSE)

writeLines(readLines(log_path))
logtree_sink_remove(h)
```

## Structured NDJSON

For anything that ends up in an aggregator rather than a terminal. One JSON
record per line, with the fields a query needs.

*Features shown:*
[output sinks](https://IvanSortino.github.io/logtree/articles/logtree.html#output-sinks),
[verbosity](https://IvanSortino.github.io/logtree/articles/logtree.html#verbosity)

```{r}
json_path <- tempfile(fileext = ".ndjson")
h <- logtree_sink_file(json_path, format = "json", threshold = "debug")

logtree_reset()

job <- function() {
  log_step("Sync accounts")
  log_debug("page size 500")
  log_info("1,204 accounts fetched")
  log_warn("12 accounts missing an email")
}

with_logging(job(), summary = FALSE)

writeLines(readLines(json_path))
logtree_sink_remove(h)
```

The console above shows no debug line -- the global threshold is still
`"info"` -- but the file has it, because this sink pinned `threshold =
"debug"` for itself. That is the point of per-sink thresholds: a verbose log
file does not force a verbose terminal.

Every record carries a `run_id`, so one run's lines can be picked out of a file
that many runs have appended to, and `ts` is ISO-8601 to the millisecond rather
than a bare epoch number:

```{r, eval = rlang::is_installed("jsonlite")}
records <- lapply(readLines(json_path), jsonlite::fromJSON)
str(records[[3]])
```

## Asserting in tests

If your package logs with logtree, test the *events*, not the rendered output.
Pattern-matching glyphs and connectors tests logtree's renderer; the memory sink
tests your code.

*Features shown:*
[testing your logging](https://IvanSortino.github.io/logtree/articles/logtree.html#testing-your-logging)

```{r}
logtree_reset()
h <- logtree_sink_memory()

import_all()

events <- logtree_sink_memory_events(h)
events[, c("level", "depth", "label", "status")]
```

`level` is the kind of event -- `"open"`, `"leaf"`, `"close"`, `"group"`,
`"group_close"` -- and `status` is its outcome.

From there the assertions are ordinary data frame work. Inside `testthat` that
looks like:

```r
test_that("a schema mismatch is logged as an error under its own file", {
  logtree_reset()
  h <- logtree_sink_memory()
  withr::defer(logtree_sink_remove(h))

  import_all()
  events <- logtree_sink_memory_events(h)

  errors <- events[events$level == "leaf" & events$status == "error", ]
  expect_equal(nrow(errors), 1L)
  expect_match(errors$label, "schema mismatch")
})
```

```{r}
# what that assertion is looking at
subset(events, level == "leaf" & status == "error",
       select = c("depth", "label"))

logtree_sink_remove(h)
```

The buffer is capped (`max = 1000` by default), so a runaway loop cannot eat
the session's memory.

## Bridging logger

An existing codebase that already calls
[logger](https://daroczig.github.io/logger/), rendered as a logtree tree without
touching any of the call sites.

*Features shown:*
[logger integration](https://IvanSortino.github.io/logtree/articles/logtree.html#logger-integration)

```{r, eval = rlang::is_installed("logger", version = "0.3.0")}
logtree_reset()
logtree_threshold("debug")

ns <- "my_app"
logtree_logger(namespace = ns)

fetch_accounts <- function() {
  log_step("Fetch accounts")
  logger::log_debug("page size 500", namespace = ns)
  logger::log_info("1,204 accounts fetched", namespace = ns)
  logger::log_warn("12 accounts missing an email", namespace = ns)
}

sync <- function() {
  log_step("Sync")
  fetch_accounts()
  logger::log_success("sync complete", namespace = ns)
}

with_logging(sync())
logtree_summary()

logtree_threshold("info")
```

```{r, echo = FALSE, eval = !rlang::is_installed("logger", version = "0.3.0")}
cat("The `logger` package (>= 0.3.0) is not installed, so this example was not run.\n")
```

The `logger::log_warn()` call elevates its enclosing logtree step exactly as
`log_warn()` would, and reaches the digest the same way -- once bridged, a
logger call is an ordinary leaf.
