---
title: "Recipes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(logtree)
```

Short answers to the questions that come up once logtree is in a real codebase.

## Logging from a top-level script

`log_step()` hangs its close on the calling function's frame. A script's top
level has no such frame -- the global environment never returns -- so the step
would stay open forever, and logtree prints a one-time nudge if you try.

Use the manual pair instead:

```{r}
logtree_reset()

id <- log_open("Load inputs")
log_info("reading 3 files")
log_success("9,412 rows")
log_close(id)

id <- log_open("Publish")
log_success("uploaded", close = TRUE)
```

The functions those blocks *call* can still use `log_step()` freely -- they have
frames. It is only the outermost, top-level step that needs opening by hand.

For error handling at that level there is no expression to wrap either, so
`with_logging(global = TRUE)` installs the handling for the rest of the script:

```r
# at the top of the script
with_logging(global = TRUE, warnings = TRUE)

id <- log_open("Nightly job")
run_everything()          # an error anywhere below is logged, then rethrown
log_close(id)
```

Global mode only acts while logtree steps are open, so a session-persistent
handler cannot swallow warnings from unrelated code later on.

## Re-running the same block interactively

At the top level a step is keyed on its own source location. Re-running the same
`log_open()` line in RStudio or Positron re-anchors to the same node instead of
nesting a level deeper each time, so an afternoon of re-evaluating the same
block does not walk the tree off the right edge of the console.

Nothing to configure -- but it is worth knowing that this is *why* re-running
behaves differently from calling the same function twice.

## Logging from a package, quietly

A package that logs with logtree should not make its own test suite noisy.
`logtree_mute()` stops every sink receiving events without unregistering any of
them, and both mute and unmute return the state they replaced, so a caller can
restore what it found rather than assuming:

```{r}
quietly <- function(code) {
  was <- logtree_mute()
  on.exit(if (!was) logtree_unmute(), add = TRUE)
  force(code)
}

logtree_reset()
quietly({
  id <- log_open("Internal work")
  log_warn("something to remember")
  log_close(id)
})

# nothing printed -- but the run was still recorded
logtree_summary()
```

That last part matters: a muted run is recorded, not discarded. The digest can
still report what went wrong, and step bookkeeping is untouched, so depth is
right the moment output comes back.

In `testthat`, scope it with `withr::defer()`:

```r
local_quiet_logtree <- function(.env = parent.frame()) {
  was <- logtree_mute()
  withr::defer(if (!was) logtree_unmute(), envir = .env)
}
```

## Keeping a log file for a scheduled job

A job nobody watches wants three things the console does not: a timestamp on
every line, a file to write to, and enough verbosity to diagnose a failure
after the fact.

```{r}
log_path <- tempfile(fileext = ".log")

h <- logtree_sink_file(
  log_path,
  format    = "text",
  timestamp = "%Y-%m-%d %H:%M:%S",
  threshold = "debug"
)

logtree_reset()

job <- function() {
  log_step("Nightly sync")
  log_debug("page size 500")
  log_info("1,204 accounts fetched")
  log_warn("12 accounts missing an email")
}

with_logging(job(), summary = FALSE)

writeLines(readLines(log_path))
logtree_sink_remove(h)
```

The console stayed at the default `"info"` and carried no timestamp; the file
pinned both for itself. That is the general shape -- `logtree_threshold()` is
the default for sinks that do not pin one, so moving it still moves them, but a
sink that has an opinion keeps it.

Use `format = "json"` instead when the file is going to an aggregator rather
than a person; every record then carries a `run_id` so one run can be picked
out of a file many runs appended to.

## Choosing a threshold per sink

A rough rule that covers most setups:

| Destination | Threshold | Why |
| --- | --- | --- |
| Console, interactive | `"info"` (default) | you are watching it happen |
| Console, CI | `"info"` | build logs are read on failure, and debug drowns the signal |
| Text file next to a job | `"debug"` | the file exists precisely for the failure you did not anticipate |
| NDJSON to an aggregator | `"debug"` | storage is cheap, and the query filters |

Remember that verbosity is a *rendering* gate only: a warning suppressed
everywhere still elevates its step's glyph and still reaches the digest.

## Sending events somewhere logtree does not support

`logtree_sink()` takes any function of one argument, called with each event.
That is the extension point -- an HTTP post, a database insert, a counter.

```{r}
levels_seen <- character(0)
h <- logtree_sink(function(event) {
  if (identical(event$kind, "leaf")) {
    levels_seen <<- c(levels_seen, event$status)
  }
})

logtree_reset()
job()
table(levels_seen)

logtree_sink_remove(h)
```

A sink that throws is skipped rather than allowed to break the fanout: the
remaining sinks still run and a warning naming the offender is raised once. Do
not rely on that as error handling, though -- a sink that talks to the network
should have its own `tryCatch()`, or a slow endpoint will pace your pipeline.

## Bridging an existing `logger` codebase

One call, near the top of the script. It registers logtree's layout, pairs it
with `logger::appender_void` so logtree does the rendering, and opens logger's
own threshold so `logtree_threshold()` becomes the single gate.

```{r, eval = rlang::is_installed("logger", version = "0.3.0")}
logtree_reset()
ns <- "my_app"
logtree_logger(namespace = ns)

process <- function() {
  log_step("Process")
  logger::log_info("reading input", namespace = ns)
  logger::log_warn("2 malformed rows skipped", namespace = ns)
}

with_logging(process(), summary = FALSE)
```

```{r, echo = FALSE, eval = !rlang::is_installed("logger", version = "0.3.0")}
cat("The `logger` package (>= 0.3.0) is not installed, so this example was not run.\n")
```

Severities map onto leaf levels: `FATAL`/`ERROR` to `log_error()`, `WARN` to
`log_warn()`, `SUCCESS` to `log_success()`, `INFO` to `log_info()`, and
`DEBUG`/`TRACE` both to `log_debug()`. Note that logger's own
`log_threshold()` and `logtree_threshold()` are two independent gates applied
on top of each other -- `logtree_logger()` opens logger's so that logtree's is
the one that matters, but if you set logger's again afterwards, both apply.

`logtree_logger()` requires `logger` 0.3.0 or newer, since that is when
`appender_void` arrived; against an older one it refuses up front with a
message naming the requirement rather than failing obscurely later.
