---
title: "API Reference"
---

Complete reference for all logtree functions.

## Core Logging

### log_step()

Open a step and automatically close it when the calling function returns.

```r
log_step(label, group_by = NULL)
```

**Arguments:**
- `label` — Step description (character)
- `group_by` — Named vector `c(name = value)` for grouping adjacent steps with the same value

**Returns:** `invisible(NULL)`. Registration of the close handler in the caller's frame happens via deferred cleanup.

**Examples:**

```{r}
#| eval: false
load_config <- function() {
  log_step("Load config")
  # ...
}
```

### log_info(), log_success(), log_warn(), log_error(), log_debug()

Log a message as a leaf attached to the current step.

```r
log_info(message)
log_success(message)
log_warn(message)
log_error(message)
log_debug(message)
```

**Arguments:**
- `message` — Message text (character)

**Returns:** `invisible(NULL)`

**Behavior:**
- `log_warn()` elevates the current step's status to warning
- `log_error()` elevates the current step's status to error
- `log_debug()` only appears if verbosity is set to "debug" via `logtree_threshold()`

## Theming

### logtree_theme()

Set or customize the active theme.

```r
logtree_theme(theme = NULL)
```

**Arguments:**
- `theme` — Either a preset name ("unicode", "ascii", "emoji") or a named list of glyph overrides

**Presets:**
- `"unicode"` — Tree connectors and status glyphs (default)
- `"ascii"` — ASCII-only, CRAN-portable
- `"emoji"` — Emoji glyphs

**Examples:**

```{r}
#| eval: false
# Switch to ASCII theme
logtree_theme("ascii")

# Custom theme
logtree_theme(list(
  running = "→",
  success = "✓",
  warning = "⚠",
  error = "✗"
))
```

## Output & Sinks

### logtree_sink_file()

Add a file sink for logging to text or JSON format.

```r
logtree_sink_file(path, format = c("text", "json"))
```

**Arguments:**
- `path` — File path (character)
- `format` — Output format: "text" for plain ASCII, "json" for NDJSON

**Returns:** `invisible(NULL)`

**Examples:**

```{r}
#| eval: false
logtree_sink_file("pipeline.log", format = "text")
logtree_sink_file("pipeline.json", format = "json")
```

## Control

### logtree_threshold()

Set the minimum verbosity level for displayed messages.

```r
logtree_threshold(level = c("debug", "info", "success", "warn", "error"))
```

**Arguments:**
- `level` — Minimum level (case-insensitive)

**Levels (most to least verbose):**
- `"debug"` — Show all messages including debug
- `"info"` — Show info and above (default)
- `"success"` — Show only success, warning, error
- `"warn"` — Show only warnings and errors
- `"error"` — Show only errors

### logtree_reset()

Clear all open steps, sinks, and reset to default theme.

```r
logtree_reset()
```

**Returns:** `invisible(NULL)`

**Use cases:** Between tests, or to recover from interrupted sessions.

### with_logging()

Install error handlers that mark all open steps as error and log the condition before rethrowing.

```r
with_logging(expr)
```

**Arguments:**
- `expr` — Expression to evaluate

**Returns:** Return value of `expr`

**Behavior:** If an error occurs during `expr`, all open steps are marked with error status, the error message is logged as a leaf, and the error is rethrown (not swallowed).

**Examples:**

```{r}
#| eval: false
result <- with_logging({
  log_step("Risky operation")
  risky_function()
  log_success("Completed")
})
```

## Logger Package Integration

### layout_logtree()

Create a log layout function for the `logger` package.

```r
layout_logtree()
```

**Returns:** A layout function suitable for `logger::log_appender()`

**Examples:**

```{r}
#| eval: false
library(logger)
library(logtree)

logger::log_appender(logtree::layout_logtree)
logger::log_info("This message goes to logtree")
```
