---
title: "Get Started"
---

## Installation

logtree is not yet on CRAN. Install from GitHub:

```{r}
#| eval: false
devtools::install_github("sortino/tree-logger")
```

Load it:

```{r}
#| eval: false
library(logtree)
```

## Basic Usage

`log_step()` opens a step and automatically closes it when the calling function returns — whether by normal return, early exit, or error.

```{r}
#| eval: false
process_data <- function() {
  log_step("Processing data")
  # Your code here
  log_info("Step completed successfully")
}

process_data()
```

Nesting happens automatically: a step opened inside another is indented one level deeper without any coordination between functions.

```{r}
#| eval: false
load_config <- function() {
  log_step("Load config")
  log_info("Reading config.yml")
  log_success("Validated 12 parameters")
}

pipeline <- function() {
  log_step("Pipeline")
  load_config()
  log_info("All done")
}

pipeline()
```

## Leaves and Statuses

Log messages at different levels as "leaves" attached to the current step:

- `log_info()` — informational message (ℹ glyph)
- `log_success()` — success message (✔ glyph)
- `log_warn()` — warning, elevates current step to warning status (⚠ glyph)
- `log_error()` — error message, elevates current step to error status (✖ glyph)
- `log_debug()` — debug message (visible only at debug verbosity level)

Step status is elevated by the most severe leaf: a step with a warning stays warning, but an error elevates it to error.

```{r}
#| eval: false
validate_data <- function() {
  log_step("Validate data")

  if (has_issues) {
    log_warn("Found 3 data anomalies")
  } else {
    log_success("All checks passed")
  }
}
```

## Grouping Steps

Group adjacent steps under a synthetic header using `group_by`:

```{r}
#| eval: false
process_file <- function(file) {
  log_step("Process", group_by = c(dataset = file))
  # Processing logic
}

for (file in c("data1.csv", "data2.csv")) {
  process_file(file)
}
```

Output groups consecutive steps with the same value under one header. When values differ or a non-grouped step appears, the group closes.

## Theming

Customize glyphs and tree connectors with built-in themes or custom definitions:

```{r}
#| eval: false
library(logtree)

# Switch to ASCII theme (portable, no emoji)
logtree_theme("ascii")

# Or use emoji theme
logtree_theme("emoji")

# Or define a custom theme
logtree_theme(
  list(
    running = "→",
    success = "•",
    warning = "!",
    error = "x"
  )
)
```

## Error Handling

logtree works correctly even when steps error. Use `with_logging()` to automatically mark all open steps as error and log the condition before rethrowing:

```{r}
#| eval: false
with_logging({
  log_step("Risky operation")
  risky_function()  # Throws error
  log_success("Completed")
})
```

If an error occurs:
1. All open steps are marked with error status
2. The error message is logged as a leaf
3. The error is rethrown (not swallowed)

Without `with_logging()`, a step whose frame exits abnormally is marked `"incomplete"` (dimmed glyph) rather than showing false success.

## Logging to Files

Send logs to a file in addition to the console:

```{r}
#| eval: false
logtree_sink_file("pipeline.log", format = "text")
# or JSON format
logtree_sink_file("pipeline.json", format = "json")

# Then log normally
log_step("Process")
```

## Integration with logger Package

Use logtree as a layout for the `logger` package:

```{r}
#| eval: false
library(logger)
library(logtree)

log_appender(appender_console_with_logtree)
```

## Verbosity Control

Control which messages appear using `logtree_threshold()`:

```{r}
#| eval: false
# Show debug messages
logtree_threshold("debug")

# Show only warnings and errors
logtree_threshold("warn")
```

Levels (from most to least verbose):
- `debug`
- `info`
- `success`
- `warn`
- `error`
