---
title: logtree
subtitle: "Live, colored tree logging for nested process execution in R"
---

logtree renders nested process execution as a live, colored tree in the console with tree connectors, status glyphs (вњ” вљ  вњ– в„№ вљ™), and elapsed time per step вЂ” while keeping nesting depth correct even when a step errors partway through.

## Quick Example

```{r}
#| eval: false
library(logtree)

demo <- function() {
  log_step("Initialize")
  log_debug("Debug: Setting up environment")
  log_info("Info: Starting process")
  log_success("Success: Initialized")

  log_step("Fetch data")
  log_info("Reading from API")
  log_success("Retrieved 1000 rows")

  log_step("Validate")
  log_warn("Warning: 5 rows have missing values")

  log_step("Transform")
  log_success("Transformation complete")
}

with_logging(demo())
```

Actual output (with all status indicators):

```
в–¶ Initialize
в”њв”Ђ вљ™ Debug: Setting up environment
в”њв”Ђ в„№ Info: Starting process
в”њв”Ђ вњ” Success: Initialized
в”њв”Ђ в–¶ Fetch data
в”‚  в”њв”Ђ в„№ Reading from API
в”‚  в”њв”Ђ вњ” Retrieved 1000 rows
в”‚  в”њв”Ђ в–¶ Validate
в”‚  в”‚  в”њв”Ђ вљ  Warning: 5 rows have missing values
в”‚  в”‚  в”њв”Ђ в–¶ Transform
в”‚  в”‚  в”‚  в”њв”Ђ вњ” Transformation complete
в”‚  в”‚  в”‚  в””в”Ђ вњ” Done  0.00s
в”‚  в”‚  в””в”Ђ вљ  Done  0.01s
в”‚  в””в”Ђ вњ” Done  0.01s
в””в”Ђ вњ” Done  0.03s
вњ” Run complete in 0.03s
```

**Status Indicators:**
- `в–¶` Running (step currently open)
- `вњ”` Success (step completed successfully)
- `вљ ` Warning (step encountered warnings)
- `вњ–` Error (step failed with an error)
- `в„№` Info message
- `вљ™` Debug message (when debug verbosity is enabled)

## Features

- **Automatic nesting**: Steps opened inside other steps are indented without coordination
- **Error resilience**: Nesting depth stays correct even when a step errors or exits early
- **Live rendering**: Watch progress in real-time as your code executes
- **Customizable themes**: Unicode, ASCII, or emoji glyphs + custom themes
- **Multiple output formats**: Console color, JSON, or file sink appenders
- **Logger integration**: Works with the `logger` package via `layout_logtree()`
- **Verbosity control**: Filter messages by level (debug, info, success, warn, error)

## Installation

```{r}
#| eval: false
# Install from GitHub (not yet on CRAN)
devtools::install_github("sortino/tree-logger")
```

## Documentation

- **[Get Started Guide](get-started.qmd)** вЂ” Basic usage and common patterns
- **[API Reference](api-reference.qmd)** вЂ” Function signatures and behavior
- **[Design Philosophy](design.qmd)** вЂ” Architecture and design decisions
- **[Source Code](https://github.com/sortino/tree-logger)** вЂ” Development on GitHub
